---
title: "Demographic Analysis"
author: "Kiet Do"
date: "2024-02-09"
output: html_document
---


# Project Scope:

The Academic Deans’ Services (ADS) is an office dedicated to supporting students in achieving their academic goals and navigating their undergraduate journey effectively. Within this project scope, we aim to understand student engagement patterns with academic support services provided by ADS. By analyzing institutional data, we seek to uncover trends and patterns that illuminate the effectiveness of one-on-one advising services offered by ADS. Additionally, we will investigate which demographic groups are being effectively reached by ADS services and identify any potential gaps or areas for improvement in outreach efforts.


```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

knitr::opts_chunk$set(echo = TRUE)



acad.year <- function(Term, TermType){
  Data <- data.frame(Term, stringsAsFactors = FALSE)

  if(TermType == "SIS"){
    Data$cent <- substr(Data$Term, 1, 1)
    Data$year <- substr(Data$Term, 2, 3)
    Data$semester <- substr(Data$Term, 4, 4)

    if (length(Data[which(Data$cent == "0"),]$cent) > 0)
      {Data[which(Data$cent == "0"),]$cent <- "19"}
    if (length(Data[which(Data$cent == "1"),]$cent) > 0)
      {Data[which(Data$cent == "1"),]$cent <- "20"}
    Data$acadYear <- paste(Data$cent, Data$year, sep = "")
    if (length(Data[which(Data$semester == "6"),]$cent) > 0)
      {Data[which(Data$semester == "6"),]$acadYear <- as.integer(Data[which(Data$semester == "6"),]$acadYear) + 1}

  }
  if(TermType == "DARS"){
    Data$acadYear <- substr(Data$Term, 1, 4)
    Data$semester <- substr(Data$Term, 5, 5)
    if (length(Data[which(Data$semester == "3"),]$semester) > 0)
      {Data[which(Data$semester == "3"),]$acadYear <- as.integer(Data[which(Data$semester == "3"),]$acadYear) + 1}

  }

  return(Data$acadYear)

}

```




```{r }
#| label: data-exploration
#| echo: false
#| message: false
#| warning: false


library(readxl)
library(tidyverse)

data <- read_excel("Random_ADS_Advisor_Notes.xlsx")
comparison.data <- read_excel("L&S_COMP.xlsx")


comparison.data <- comparison.data |>
        filter(TERM > 1136) |>
        mutate(AcadYear = acad.year(TERM, "SIS"))


comparison.gender  <- comparison.data |>
          group_by(AcadYear, GENDER) |>
          summarize(n = sum(N)) |>
          ungroup() |>
          group_by(AcadYear) |>
          mutate(`L&S_prop` = n/sum(n)) |>
          select(-n) |>
          rename(FirstAcadYear = AcadYear)

comparison.minority <- comparison.data |>
          group_by(AcadYear, MINORITY) |>
          summarize(n = sum(N)) |>
          ungroup() |>
          group_by(AcadYear) |>
          mutate(`L&S_prop` = n / sum(n)) |>
          select(-n) |>
          rename(FirstAcadYear = AcadYear)

comparison.residency <- comparison.data |>
         mutate(Residency_Status = case_when(
    RACE == "International" ~ "International",
    TUITION_RESIDENCY == "N" ~ "Out of State",
    TUITION_RESIDENCY == "R" ~ "WI resident",
    TUITION_RESIDENCY == "S" ~ "MN compact"
  )) |>
        group_by(AcadYear, Residency_Status) |>
        summarize(n = sum(N)) |>
        ungroup() |>
        group_by(AcadYear) |>
        mutate(`L&S_prop` = n / sum(n)) |>
        select(-n) |>
        rename(FirstAcadYear = AcadYear)

comparison.targeted_minority <- comparison.data |>
        group_by(AcadYear, TARGETED_MINORITY) |>
        summarize(n = sum(N)) |>
        ungroup() |>
        group_by(AcadYear) |>
        mutate(`L&S_prop` = n / sum(n)) |>
        select(-n) |>
        rename(FirstAcadYear = AcadYear)

data <- data |>
      replace_na(list(MINORITY = "N", 
                      TARGETED_MINORITY = "N")) |>
    mutate(FirstAcadYear = acad.year(FIRST_SEMESTER_ENROLLED,
                                "SIS"))


data <- data |>
  mutate(Residency_Status = case_when(
    ETHNIC_REPORTING_CODE == 10 ~ "International",
    TUITION_RESIDENCY_AT_ENTRANCE == "N" ~ "Out of State",
    TUITION_RESIDENCY_AT_ENTRANCE == "R" ~ "WI resident",
    TUITION_RESIDENCY_AT_ENTRANCE == "S" ~ "MN compact"
  ))


data$CONTACT_DATE <- as.Date(data$CONTACT_DATE)
data$contact_year <- as.integer(format(data$CONTACT_DATE, "%Y"))

```



```{r}
#| echo: false
#| message: false
#| warning: false


# Gender Trends
 data |>
  filter(GENDER != "U", FirstAcadYear > 2010) |>
  group_by(FirstAcadYear, GENDER) |>
  summarize(n = n_distinct(Random_ID)) |>
  group_by(FirstAcadYear) |>
  mutate(prop = n / sum(n)) |>
  filter(GENDER == "F") |>
  inner_join(comparison.gender, by = c("FirstAcadYear", "GENDER")) |>
  ggplot(aes(x = FirstAcadYear, y = prop, group = GENDER)) +
  geom_line() +
  geom_line(aes(y = `L&S_prop`), linetype = 2) + 
  scale_y_continuous(limits = c(0,.7), 
                     breaks = c(0,.1,.2,.3,.4,.5,.6,.7),
                     labels = scales::percent) +
  labs(title = "Gender Trends by First Academic Year",
       subtitle = "Dotted line is L&S entrants in that year.",
       x = "Academic Year",
       y = "%") +
   theme_minimal()


```
Gender ratio has been stable for the last 10 years

```{r}
#| echo: false
#| message: false
#| warning: false


# Minority Trends
data |>
  filter(FirstAcadYear > "2010", ETHNIC_REPORTING_DESCR != "International") |>
  group_by(FirstAcadYear, MINORITY) |>
  summarize(n = n_distinct(Random_ID)) |>
  group_by(FirstAcadYear) |>
  mutate(prop = n / sum(n)) |>
  filter(MINORITY == "Y") |>
  inner_join(comparison.minority, by = c("FirstAcadYear", "MINORITY")) |>
  ggplot(aes(x = FirstAcadYear, y = prop, group = MINORITY)) +
  geom_line() +
  geom_line(aes(y = `L&S_prop`), linetype = 2) +
  scale_y_continuous(limits = c(0, 0.5), 
                     breaks = seq(0, 0.5, 0.05),
                     labels = scales::percent) +
  labs(title = "Minority Trends by First Academic Year",
       subtitle = "Dotted line is L&S entrants in that year.",
       x = "Academic Year",
       y = "%") +
  theme_minimal()


```

```{r}
#| echo: false
#| message: false
#| warning: false


# Residency Trends
residency_trends <- data |>
  filter(FirstAcadYear > "2010") |>
  group_by(FirstAcadYear, Residency_Status) |>
  summarize(n = n_distinct(Random_ID)) |>
  group_by(FirstAcadYear) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = FirstAcadYear, y = prop, fill = Residency_Status)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Residency Trends by First Academic Year",
       y = "Proportion",
       fill = "Residency")

residency_trends
```

```{r}
#| echo: false
#| message: false
#| warning: false


# Residency Trends
data |>
  filter(FirstAcadYear > "2010") |>
  group_by(FirstAcadYear, Residency_Status) |>
  summarize(n = n_distinct(Random_ID)) |>
  group_by(FirstAcadYear) |>
  mutate(ADS_prop = n / sum(n)) |>
  #filter(Residency_Status == "International") |>
  inner_join(comparison.residency, by = c("FirstAcadYear" , "Residency_Status" )) |>
  select(-n) |>
  pivot_longer(cols = c("ADS_prop", "L&S_prop"), names_to = "Group", values_to = "ADS_prop") |>
  
  ggplot(aes(x = FirstAcadYear, y = ADS_prop, group = Group)) +
  geom_line(aes(y = ADS_prop, linetype = Group)) +
  #scale_y_continuous(limits = c(0, 0.5), 
  #                   breaks = seq(0, 0.5, 0.05),
   #                  labels = scales::percent) +
  facet_wrap(~ Residency_Status) +
  labs(title = "Residency Trends by First Academic Year",
       x = "Academic Year",
       y = "%") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
#| echo: false
#| message: false
#| warning: false


data |>
  filter(FirstAcadYear > "2010", ETHNIC_REPORTING_DESCR != "International") |>
  group_by(FirstAcadYear, TARGETED_MINORITY) |>
  summarize(n = n_distinct(Random_ID)) |>
  group_by(FirstAcadYear) |>
  mutate(prop = n / sum(n)) |>
  filter(TARGETED_MINORITY == "Y") |>
  inner_join(comparison.targeted_minority, by = c("FirstAcadYear", "TARGETED_MINORITY")) |>
  ggplot(aes(x = FirstAcadYear, y = prop, group = TARGETED_MINORITY)) +
  geom_line() +
  geom_line(aes(y = `L&S_prop`), linetype = 2) +
  scale_y_continuous(limits = c(0, 0.3), 
                     breaks = seq(0, 0.3, 0.05),
                     labels = scales::percent) +
  labs(title = "Targeted Minority Trends by First Academic Year",
       subtitle = "Dotted line is L&S entrants in that year.",
       x = "Academic Year",
       y = "%") +
  theme_minimal()
```

```{r}
#| echo: false
#| message: false
#| warning: false

# Ethnicity Breakdown for Minorities
ethnicity_trends <- data |>
  filter(FirstAcadYear >= "2010", ETHNIC_REPORTING_DESCR != "International", MINORITY == "Y") |>
  group_by(FirstAcadYear, ETHNIC_REPORTING_DESCR) |>
  summarize(n = n_distinct(Random_ID)) |>
  group_by(FirstAcadYear) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = FirstAcadYear, y = prop, fill = ETHNIC_REPORTING_DESCR)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Ethnicity Breakdown for Minorities by First Academic Year",
       y = "Proportion",
       fill = "Ethnicity")

ethnicity_trends
```

```{r}
#| echo: false
#| message: false
#| warning: false


reason_counts <- data |>
  count(CONTACT_REASON1_NAME) |>
  arrange(desc(n))

# Identify top 5 reasons
top_5_reasons <- head(reason_counts, 5)

# Summarize other reasons
other_reasons <- summarise(reason_counts |> filter(!(CONTACT_REASON1_NAME %in% top_5_reasons$CONTACT_REASON1_NAME)), 
                           CONTACT_REASON1_NAME = "Others", n = sum(n))

# Combine top 5 reasons and "Others"
reason_data <- bind_rows(top_5_reasons, other_reasons)

# Plotting
reason_data |>
  ggplot(aes(x = "", y = n, fill = CONTACT_REASON1_NAME)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Top 5 Reasons for Contact",
       fill = "Reason") +
  theme_void() +
  theme(legend.position = "right")
```

```{r}
#| echo: false
#| message: false
#| warning: false

# Filter data for the last 10 years
data_last_10_years <- data |>
  filter(contact_year > 2010)

# Group data by contact reason and academic year, then count occurrences
contact_reason_counts <- data_last_10_years |>
  group_by(contact_year, CONTACT_REASON1_NAME) |>
  summarize(n = n()) |>
  arrange(contact_year, desc(n))

# Identify the top 5 contact reasons
top_5_reasons <- contact_reason_counts |>
  group_by(CONTACT_REASON1_NAME) |>
  summarize(total_occurrences = sum(n)) |>
  top_n(5, total_occurrences) |>
  select(CONTACT_REASON1_NAME)

# Filter the data to include only the top 5 contact reasons
top_5_contact_reason_counts <- contact_reason_counts |>
  filter(CONTACT_REASON1_NAME %in% top_5_reasons$CONTACT_REASON1_NAME)

# Plot bar chart for the top 5 contact reasons
ggplot(top_5_contact_reason_counts, aes(x = factor(contact_year), y = n, fill = CONTACT_REASON1_NAME)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Popularity of Top 5 Contact Reasons Over the Last 10 Years",
       x = "Academic Year",
       y = "Number of Occurrences",
       fill = "Contact Reason") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)

# Group data by year and count the number of contacts
total_contacts <- data %>%
  group_by(contact_year) %>%
  summarise(total_contacts = length(Random_ID))

# Plotting
ggplot(total_contacts, aes(x = contact_year, y = total_contacts)) +
  geom_line() +
  geom_point() +
  labs(title = "Total Number of Contacts Each Year",
       x = "Year",
       y = "Total Number of Contacts") +
  theme_minimal()

```



```{r}
#| echo: false
#| message: false
#| warning: false

# Filter data for the last 10 years
data_last_10_years <- data |>
  filter(contact_year >= 2012 & contact_year <= 2023)

# Group data by contact reason, residency status, and contact year, then count occurrences
contact_reason_counts_by_residency <- data_last_10_years |>
  filter(Residency_Status != "N/A") |>
  group_by(Residency_Status, CONTACT_REASON1_NAME, contact_year) |>
  summarize(n = n()) |>
  arrange(Residency_Status, contact_year, desc(n))

contact_reason_counts_by_residency$contact_year <- as.integer(contact_reason_counts_by_residency$contact_year)

# Identify the top 5 contact reasons for each residency status
top_5_reasons_by_residency <- contact_reason_counts_by_residency |>
  group_by(Residency_Status, CONTACT_REASON1_NAME) |>
  summarize(total_occurrences = sum(n)) |>
  top_n(5, total_occurrences) |>
  select(Residency_Status, CONTACT_REASON1_NAME)

# Calculate the total number of occurrences for each residency group and each year
total_counts <- contact_reason_counts_by_residency |>
  group_by(Residency_Status, contact_year) |>
  summarize(total_occurrences = sum(n))

# Join the total counts with the contact_reason_counts_by_residency dataframe
contact_reason_counts_by_residency <- contact_reason_counts_by_residency |>
  left_join(total_counts, by = c("Residency_Status", "contact_year"))

# Calculate the percentage of each contact reason relative to the total for each residency group and each year
contact_reason_counts_by_residency <- contact_reason_counts_by_residency |>
  mutate(percentage = (n / total_occurrences) * 100)

# Filter the data to include only the top 5 contact reasons for each residency status
top_5_contact_reason_counts_by_residency <- contact_reason_counts_by_residency |>
  filter(CONTACT_REASON1_NAME %in% top_5_reasons_by_residency$CONTACT_REASON1_NAME)

# Plot bar chart for the top 5 contact reasons for each residency status
ggplot(top_5_contact_reason_counts_by_residency, aes(x = contact_year, y = percentage, fill = CONTACT_REASON1_NAME)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Top 5 Contact Reasons by Residency Group Over the Last 10 Years",
       x = "Contact Year",
       y = "Percentage of Total Occurrences",
       fill = "Contact Reason") +
  facet_wrap(~ Residency_Status, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
#| echo: false
#| message: false
#| warning: false

data_last_10_years <- data |>
  filter(contact_year == c(2020, 2021, 2022))

# Filter data for international students
international_students <- data_last_10_years |>
  filter(Residency_Status == "International")

# Group data by contact reason for international students and count occurrences
contact_reason_counts_international <- international_students |>
  group_by(CONTACT_REASON1_NAME) |>
  summarize(n = n()) |>
  arrange(desc(n))

# Plot bar chart for contact reasons of international students
ggplot(contact_reason_counts_international, aes(x = reorder(CONTACT_REASON1_NAME, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Contact Reasons of International Students",
       subtitle = "During 2020, 2021, 2022 school years",
       x = "Contact Reason",
       y = "Number of Occurrences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title = element_text(hjust = 0.5))

```


```{r}
#| echo: false
#| message: false
#| warning: false

# Filter data for the last 10 years
data_last_10_years <- data |>
  filter(contact_year == c(2020, 2021, 2022))

# Filter data for international students
international_students <- data_last_10_years |>
  filter(Residency_Status == "WI resident")

# Group data by contact reason for international students and count occurrences
contact_reason_counts_international <- international_students |>
  group_by(CONTACT_REASON1_NAME) |>
  summarize(n = n()) |>
  arrange(desc(n))

# Plot bar chart for contact reasons of international students
ggplot(contact_reason_counts_international, aes(x = reorder(CONTACT_REASON1_NAME, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Contact Reasons of WI resident Students",
       subtitle = "During 2020, 2021, 2022 school years",
       x = "Contact Reason",
       y = "Number of Occurrences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title = element_text(hjust = 0.5))

```

```{r}
#| echo: false
#| message: false
#| warning: false

# Filter data for the last 10 years
data_last_10_years <- data |>
  filter(contact_year == c(2020, 2021, 2022))

# Filter data for international students
international_students <- data_last_10_years |>
  filter(Residency_Status == "Out of State")

# Group data by contact reason for international students and count occurrences
contact_reason_counts_international <- international_students |>
  group_by(CONTACT_REASON1_NAME) |>
  summarize(n = n()) |>
  arrange(desc(n))

# Plot bar chart for contact reasons of international students
ggplot(contact_reason_counts_international, aes(x = reorder(CONTACT_REASON1_NAME, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Contact Reasons of Out of State Students",
       subtitle = "During 2020, 2021, 2022 school years",
       x = "Contact Reason",
       y = "Number of Occurrences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title = element_text(hjust = 0.5))

```


```{r}
#| echo: false
#| message: false
#| warning: false

# Filter data for the last 10 years
data_last_10_years <- data |>
  filter(contact_year == c(2020, 2021, 2022))

# Filter data for international students
international_students <- data_last_10_years |>
  filter(Residency_Status == "MN compact")

# Group data by contact reason for international students and count occurrences
contact_reason_counts_international <- international_students |>
  group_by(CONTACT_REASON1_NAME) |>
  summarize(n = n()) |>
  arrange(desc(n))

# Plot bar chart for contact reasons of international students
ggplot(contact_reason_counts_international, aes(x = reorder(CONTACT_REASON1_NAME, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Contact Reasons of MN Compact Students",
       subtitle = "During 2020, 2021, 2022 school years",
       x = "Contact Reason",
       y = "Number of Occurrences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.title = element_text(hjust = 0.5))

```
